---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

I unzip and load the data.

```{r}
unzip("activity.zip")
df <- read.csv("activity.csv")
summary(df)
```

The interval has a time format:

```{r}
unique(df$interval)
```

I combine the date and the interval into a datetime.

```{r}
df$datetime <- as.POSIXct(paste(df$date, sprintf("%.4d", df$interval)), format="%Y-%m-%d %H%M")
head(df$datetime)
```

I parse the date string into a date value. 

```{r}
df$date <- as.Date(df$date, format="%Y-%m-%d")
head(df$date)
```

I parse the interval into a datetime where I will only use the time. (Note that todays date is going to be used, but we will ignore this part)

```{r}
df$time <- as.POSIXct(sprintf("%.4d", df$interval), format="%H%M")
head(df$time)
```


## What is mean total number of steps taken per day?

I keep the part of the data where the steps are not missing.

```{r}
df.not.na <- df[!is.na(df$steps),]
day.total <- aggregate(df.not.na$steps, list(df.not.na$date), sum)
names(day.total) <- c("date", "steps")

meanTotalSteps <- mean(day.total$steps)
medianTotalSteps <- median(day.total$steps)
```

The total number of steps per day has

- mean equal to **`r sprintf("%.1f", meanTotalSteps)`**, and
- median equal to **`r medianTotalSteps`**.

I make a histogram of the total number of steps pre day, and draw the mean and the median.

```{r}
library(ggplot2)
h <- ggplot(day.total, aes(x=steps)) + 
     geom_histogram(binwidth=1000, colour="black", fill="red") + 
     xlab("Total number of steps per day") +
     geom_vline(xintercept=meanTotalSteps, color="blue", size=1.2) +
     geom_vline(xintercept=medianTotalSteps, color="green", linetype="dashed", size=1.2) + 
     geom_text(aes(x=meanTotalSteps, label="\nMean", y=7), 
               colour="blue", angle=90, text=element_text(size=10)) +
     geom_text(aes(x=medianTotalSteps, label="Median\n", y=7), 
               colour="green", angle=90, text=element_text(size=10))
h
```


## What is the average daily activity pattern?

I average the number of steps for each 5-minute interval accross all days.

```{r}
interval.average <- aggregate(df.not.na$steps, list(df.not.na$time), mean)
names(interval.average) <- c("time", "steps")
```

I find the time of the interval with the maximum number of steps.

```{r}
maxStepInterval <- interval.average[which.max(interval.average$steps),]$time
maxStepTime <- strftime(maxStepInterval, format="%H:%M")
```

On average across all the days in the dataset, the maximum number of steps are in the 5-minute interval at **`r maxStepTime`**.

I make a time series plot of the average number of steps for each 5-minute interval of the day.

```{r}
library(scales)

p <- ggplot(interval.average, aes(x=time, y=steps)) + geom_line() +
     xlab("5-minute interval") +
     ylab("Average number of steps") +
     scale_x_datetime(breaks=("2 hour"), 
                      labels=date_format("%H:%M")) +
     geom_vline(xintercept=as.numeric(maxStepInterval), color="red", linetype="dashed", size=0.75) +
     geom_text(aes(x=maxStepInterval, label=paste(maxStepTime, "\n"), y=25), 
               colour="red", angle=90, text=element_text(size=12))
     
p
```

## Imputing missing values

The number of missing values for each column is:

```{r}
sapply(names(df), function(c) sum(is.na(df[,c])))
```

I add the day of the week to the data frame.

```{r}
df$day <- weekdays(as.Date(df$date))
head(df$day)
```

I add a column that contains the date of the week and the 5-minute interval combined into one field.

```{r}
df$day.interval <- as.factor(paste(df$day, sprintf("%.4d", df$interval)))
head(df$day.interval)
```

I create a new data frame and fill in the missing steps with the average steps accross all weeks, for the same interval and day of the week.

```{r}
df.imputed <- transform(df, 
                        steps = ifelse(is.na(steps), 
                                       ave(df$steps, 
                                           df$day.interval, 
                                           FUN=function(x) mean(x, na.rm=TRUE)), 
                                       steps))
```

I double check that there are no missing values in the imputed data frame.

```{r}
sapply(names(df.imputed), function(c) sum(is.na(df.imputed[,c])))
```

```{r}
day.total2 <- aggregate(df.imputed$steps, list(df.imputed$date), sum)
names(day.total2) <- c("date", "steps")
meanTotalSteps2 <- mean(day.total2$steps)
medianTotalSteps2 <- median(day.total2$steps)
```

The total number of steps per day in the imputed data has

- mean equal to **`r sprintf("%.1f", meanTotalSteps2)`**, and
- median equal to **`r sprintf("%.1f", medianTotalSteps2)`**.

The change in these values compared to the data with missing values is

- mean change: **`r sprintf("%.1f", meanTotalSteps2-meanTotalSteps)`**, and
- median change: **`r sprintf("%.1f", medianTotalSteps2-medianTotalSteps)`**.

Both the mean and median have increased. The median has increased 
**`r sprintf("%.1f", (medianTotalSteps2 - medianTotalSteps)/(meanTotalSteps2-meanTotalSteps))`**
times more compared to the increase of the mean.

I make a histogram of the total number of steps per day, and draw the mean and the median, this time for the imputed data.

```{r}
h2 <- ggplot(day.total2, aes(x=steps)) + 
      geom_histogram(binwidth=1000, colour="black", fill="red") + 
      xlab("Total number of steps per day") +
      geom_vline(xintercept=meanTotalSteps2, color="blue", size=1.2) +
      geom_vline(xintercept=medianTotalSteps2, color="green", linetype="dashed", size=1.2) + 
      geom_text(aes(x=meanTotalSteps2, label="Mean\n", y=4), 
                colour="blue", angle=90, text=element_text(size=10)) +
      geom_text(aes(x=medianTotalSteps2, label="\nMedian", y=4), 
                colour="green", angle=90, text=element_text(size=10))
h2
```

## Are there differences in activity patterns between weekdays and weekends?

I create a factor variable that tells whether the sample is during the weekend or a weekday.
```{r}
df.imputed$weekday = as.factor(ifelse(df.imputed$day == "Saturday" | df.imputed$day == "Sunday", 
                                      "weekend", "weekday"))
```

I take the the mean number of steps for each interval during the week and during the weekend separately.

```{r}
weekday.interval.average <- aggregate(steps ~ time + weekday, data=df.imputed, mean)
```

I plot the average steps during the week and during the weekend.

```{r}
p2 <- ggplot(weekday.interval.average, aes(x=time,y=steps)) + 
      geom_line() +
      xlab("5-minute interval") +
      ylab("Average number of steps") +
      facet_grid(weekday ~ .) +
      scale_x_datetime(breaks=("2 hour"), 
                       labels=date_format("%H:%M"))
p2

```


